---
title: "Exploring a Large Hospital Data Set"
author: "Sabine Umuhire"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-title: "Table of Contents"
    theme: cosmo
    code-fold: true
    code-tools: true
    df-print: paged
execute:
  warning: false
  message: false
  cache: false
---

# Introduction

This report documents an exploratory analysis of a synthetic hospital database containing information on patients, clinical encounters, diagnoses, medications, departments, and providers. The goal is to characterize the data in preparation for integration into a larger analytics database. All analyses were performed in R using tidyverse tools.

```{r setup}
library(tidyverse)
library(lubridate)
library(knitr)
library(kableExtra)
library(scales)
```

---

# Part 1

## A) Data Import

The eight CSV files are imported with `readr::read_csv()`. Missing values encoded as
`"NULL"` are converted to `NA` at import via the `na` argument. Date-time columns are
parsed with `lubridate` after import.

```{r import}
data_path <- "Hospital Data Exploration/"

patients <- read_csv(
  paste0(data_path, "patients.csv"),
  na = c("", "NA", "NULL"),
  show_col_types = FALSE
)

encounters <- read_csv(
  paste0(data_path, "encounters.csv"),
  na = c("", "NA", "NULL"),
  show_col_types = FALSE
)

medications <- read_csv(
  paste0(data_path, "medications.csv"),
  na = c("", "NA", "NULL"),
  show_col_types = FALSE
)

medication_types <- read_csv(
  paste0(data_path, "medication_types.csv"),
  na = c("", "NA", "NULL"),
  show_col_types = FALSE
)

diseases <- read_csv(
  paste0(data_path, "diseases.csv"),
  na = c("", "NA", "NULL"),
  show_col_types = FALSE
)

disease_types <- read_csv(
  paste0(data_path, "disease_types.csv"),
  na = c("", "NA", "NULL"),
  show_col_types = FALSE
)

departments <- read_csv(
  paste0(data_path, "departments.csv"),
  na = c("", "NA", "NULL"),
  show_col_types = FALSE
)

providers <- read_csv(
  paste0(data_path, "providers.csv"),
  na = c("", "NA", "NULL"),
  show_col_types = FALSE
)
```

```{r parse-dates}
# Convert date/datetime columns using lubridate
patients <- patients |>
  mutate(birth_date = ymd(birth_date))

encounters <- encounters |>
  mutate(
    admit_date    = ymd_hms(admit_date),
    discharge_date = ymd_hms(discharge_date)
  )

medications <- medications |>
  mutate(date = ymd_hms(date))

diseases <- diseases |>
  mutate(date = ymd_hms(date))
```

All eight tables are now loaded as tidy data frames with proper date/datetime types and
`NA` for missing values.

---

## B) Data Summary

### B1–B8: Record Counts

```{r counts}
tibble(
  Object            = c("Patients", "Encounters", "Medication types",
                        "Medications (prescribed)", "Disease types",
                        "Diseases (diagnosed)", "Departments", "Providers"),
  Count = c(
    nrow(patients),
    nrow(encounters),
    nrow(medication_types),
    nrow(medications),
    nrow(disease_types),
    nrow(diseases),
    nrow(departments),
    nrow(providers)
  )
) |>
  mutate(Count = format(Count, big.mark = ",")) |>
  kable(caption = "Record counts across all tables") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

### B9: Patients Stratified by Sex/Gender, Race/Ethnicity, and Marital Status

The table below shows the number of patients at every combination of gender, race, and
marital status. Rows with zero patients are omitted for readability.

```{r cross-tab}
patients |>
  mutate(
    gender         = replace_na(gender, "UNKNOWN"),
    race           = replace_na(race, "UNKNOWN"),
    marital_status = replace_na(marital_status, "UNKNOWN")
  ) |>
  count(gender, race, marital_status, name = "n_patients") |>
  arrange(gender, race, marital_status) |>
  rename(
    Gender         = gender,
    Race           = race,
    `Marital Status` = marital_status,
    `N Patients`   = n_patients
  ) |>
  kable(
    caption = "Number of patients by gender, race/ethnicity, and marital status"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  scroll_box(height = "400px")
```

### B10: Top 10 Most Common Medications, Diseases, Departments, and Providers

**Top 10 Medications (by prescriptions)**

```{r top-meds}
medications |>
  left_join(medication_types, by = "medication_id") |>
  count(medication_id, medication_name, sort = TRUE) |>
  slice_head(n = 10) |>
  rename(
    `Medication ID`   = medication_id,
    `Medication Name` = medication_name,
    `Prescription Count` = n
  ) |>
  mutate(`Prescription Count` = format(`Prescription Count`, big.mark = ",")) |>
  kable(caption = "Ten most commonly prescribed medications") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

**Top 10 Diseases (by diagnoses)**

```{r top-diseases}
diseases |>
  left_join(disease_types, by = "disease_id") |>
  count(disease_id, name, sort = TRUE) |>
  slice_head(n = 10) |>
  rename(
    `Disease ID`       = disease_id,
    `Disease Name`     = name,
    `Diagnosis Count`  = n
  ) |>
  mutate(`Diagnosis Count` = format(`Diagnosis Count`, big.mark = ",")) |>
  kable(caption = "Ten most commonly diagnosed diseases") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

**Top 10 Departments (by encounters)**

```{r top-depts}
encounters |>
  left_join(departments, by = "department_id") |>
  count(department_id, department_name, sort = TRUE) |>
  slice_head(n = 10) |>
  rename(
    `Department ID`   = department_id,
    `Department Name` = department_name,
    `Encounter Count` = n
  ) |>
  mutate(`Encounter Count` = format(`Encounter Count`, big.mark = ",")) |>
  kable(caption = "Ten departments with the most encounters") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

**Top 10 Providers (by encounters, counted by ID)**

Per the instructions, providers are counted by ID (not name), in case multiple providers
share a name.

```{r top-providers}
encounters |>
  left_join(
    providers |>
      mutate(full_name = paste(first_name, last_name)),
    by = "provider_id"
  ) |>
  count(provider_id, full_name, sort = TRUE) |>
  slice_head(n = 10) |>
  rename(
    `Provider ID`     = provider_id,
    `Provider Name`   = full_name,
    `Encounter Count` = n
  ) |>
  mutate(`Encounter Count` = format(`Encounter Count`, big.mark = ",")) |>
  kable(caption = "Ten providers with the most encounters (counted by provider ID)") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

---

## C) Data Manipulation

### C1: Mean BMI by Sex/Gender; by Race/Ethnicity; and by Both

BMI is calculated from height (inches) and weight (pounds) recorded at each encounter.
The formula is: BMI = (weight_lbs / height_in²) × 703. All encounter-level
height/weight observations are used (no deduplication by patient here; see Part 2 for
the latest-only analysis).

```{r bmi-calc}
# Join encounters with patients to bring in gender and race
enc_pat <- encounters |>
  filter(!is.na(height), !is.na(weight), height > 0, weight > 0) |>
  mutate(bmi = (weight / height^2) * 703) |>
  left_join(patients |> select(patient_id, gender, race), by = "patient_id") |>
  mutate(
    gender = replace_na(gender, "UNKNOWN"),
    race   = replace_na(race, "UNKNOWN")
  )
```

**By sex/gender:**

```{r bmi-gender}
enc_pat |>
  group_by(gender) |>
  summarise(
    n          = n(),
    mean_bmi   = round(mean(bmi, na.rm = TRUE), 2),
    median_bmi = round(median(bmi, na.rm = TRUE), 2),
    sd_bmi     = round(sd(bmi, na.rm = TRUE), 2)
  ) |>
  rename(Gender = gender, N = n, `Mean BMI` = mean_bmi,
         `Median BMI` = median_bmi, `SD BMI` = sd_bmi) |>
  kable(caption = "Mean BMI by sex/gender (all encounter observations)") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

**By race/ethnicity:**

```{r bmi-race}
enc_pat |>
  group_by(race) |>
  summarise(
    n          = n(),
    mean_bmi   = round(mean(bmi, na.rm = TRUE), 2),
    median_bmi = round(median(bmi, na.rm = TRUE), 2),
    sd_bmi     = round(sd(bmi, na.rm = TRUE), 2)
  ) |>
  arrange(desc(n)) |>
  rename(Race = race, N = n, `Mean BMI` = mean_bmi,
         `Median BMI` = median_bmi, `SD BMI` = sd_bmi) |>
  kable(caption = "Mean BMI by race/ethnicity (all encounter observations)") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

**By sex/gender and race/ethnicity simultaneously:**

```{r bmi-gender-race}
enc_pat |>
  group_by(gender, race) |>
  summarise(
    n        = n(),
    mean_bmi = round(mean(bmi, na.rm = TRUE), 2),
    .groups  = "drop"
  ) |>
  arrange(gender, race) |>
  rename(Gender = gender, Race = race, N = n, `Mean BMI` = mean_bmi) |>
  kable(caption = "Mean BMI by sex/gender and race/ethnicity") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  scroll_box(height = "400px")
```

### C2: Mean Length of Stay (LOS > 1 day)

LOS is calculated as the difference in days between discharge and admit date. Encounters
with a missing discharge date or a stay of one day or less are excluded.

```{r los-calc}
enc_los <- encounters |>
  filter(!is.na(discharge_date)) |>
  mutate(los_days = as.numeric(difftime(discharge_date, admit_date, units = "days"))) |>
  filter(los_days > 1) |>
  left_join(patients |> select(patient_id, gender), by = "patient_id") |>
  left_join(departments, by = "department_id") |>
  mutate(gender = replace_na(gender, "UNKNOWN"))
```

**Overall mean LOS:**

```{r los-overall}
enc_los |>
  summarise(
    n_encounters = n(),
    mean_los     = round(mean(los_days), 2),
    median_los   = round(median(los_days), 2),
    sd_los       = round(sd(los_days), 2)
  ) |>
  rename(`N Encounters` = n_encounters, `Mean LOS (days)` = mean_los,
         `Median LOS (days)` = median_los, `SD LOS (days)` = sd_los) |>
  kable(caption = "Overall mean LOS (encounters > 1 day only)") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

**By sex/gender:**

```{r los-gender}
enc_los |>
  group_by(gender) |>
  summarise(
    n        = n(),
    mean_los = round(mean(los_days), 2),
    .groups  = "drop"
  ) |>
  rename(Gender = gender, N = n, `Mean LOS (days)` = mean_los) |>
  kable(caption = "Mean LOS by sex/gender") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

**By department:**

```{r los-dept}
enc_los |>
  group_by(department_name) |>
  summarise(
    n        = n(),
    mean_los = round(mean(los_days), 2),
    .groups  = "drop"
  ) |>
  arrange(desc(mean_los)) |>
  rename(`Department` = department_name, N = n, `Mean LOS (days)` = mean_los) |>
  kable(caption = "Mean LOS by department (sorted by mean LOS, descending)") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  scroll_box(height = "400px")
```

**By sex/gender and department simultaneously:**

```{r los-gender-dept}
enc_los |>
  group_by(gender, department_name) |>
  summarise(
    n        = n(),
    mean_los = round(mean(los_days), 2),
    .groups  = "drop"
  ) |>
  arrange(gender, desc(mean_los)) |>
  rename(Gender = gender, Department = department_name,
         N = n, `Mean LOS (days)` = mean_los) |>
  kable(caption = "Mean LOS by sex/gender and department") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  scroll_box(height = "500px")
```

---

## D) Data Visualization

### D1: Histogram of Patient Age at Encounter

```{r hist-age, fig.width=8, fig.height=5}
encounters |>
  filter(!is.na(age)) |>
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white", alpha = 0.85) +
  scale_x_continuous(breaks = seq(0, 120, 10)) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title    = "Distribution of Patient Age at Encounter",
    subtitle = "All encounters with a recorded age value",
    x        = "Age (years)",
    y        = "Number of Encounters"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

### D2: Histogram of Number of Encounters per Patient

```{r hist-enc-count, fig.width=8, fig.height=5}
enc_counts <- encounters |>
  count(patient_id, name = "n_encounters")

# Check the distribution to set a sensible x-axis cap
x_cap <- quantile(enc_counts$n_encounters, 0.99)  # show 99th percentile

enc_counts |>
  filter(n_encounters <= x_cap) |>
  ggplot(aes(x = n_encounters)) +
  geom_histogram(binwidth = 1, fill = "salmon", color = "white", alpha = 0.85) +
  scale_x_continuous(breaks = seq(0, x_cap, by = 5)) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title    = "Distribution of Number of Encounters per Patient",
    subtitle = paste0("Showing patients with up to ", x_cap,
                      " encounters (99th percentile); extreme outliers excluded"),
    x        = "Number of Encounters",
    y        = "Number of Patients"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

### D3: Scatterplot of BMI by Age, Colored by Sex/Gender

Due to the large number of observations, a random 5% sample is plotted for readability.

```{r scatter-bmi-age, fig.width=9, fig.height=6}
set.seed(42)

gender_colors <- c(
  "F" = "red",
  "M" = "steelblue",
  "A" = "forestgreen",
  "E" = "orange",
  "U" = "purple"
)

enc_pat |>
  filter(!is.na(bmi), bmi > 0, bmi < 100, !is.na(age)) |>
  slice_sample(prop = 0.05) |>
  ggplot(aes(x = age, y = bmi, color = gender)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) +
  scale_color_manual(
    values = gender_colors,
    labels = c(F = "Female", M = "Male", A = "Trans Male",
               E = "Trans Female", U = "Unknown"),
    name   = "Sex/Gender"
  ) +
  scale_x_continuous(breaks = seq(0, 120, 10)) +
  labs(
    title    = "BMI by Age at Encounter, by Sex/Gender",
    subtitle = "5% random sample of encounter records with valid BMI and age",
    x        = "Age (years)",
    y        = "BMI (kg/m²)"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```

### D4: Faceted Scatterplots of BMI by Age – by Sex/Gender and Race/Ethnicity

Panels are created for each combination of gender and race. Smaller subgroups (fewer
than 50 valid BMI observations) are excluded to avoid over-sparse panels.

```{r facet-bmi, fig.width=14, fig.height=10}
set.seed(42)

# Identify gender × race combos with at least 50 valid BMI records
keep_combos <- enc_pat |>
  filter(!is.na(bmi), bmi > 0, bmi < 100, !is.na(age)) |>
  count(gender, race) |>
  filter(n >= 50)

enc_pat |>
  filter(!is.na(bmi), bmi > 0, bmi < 100, !is.na(age)) |>
  inner_join(keep_combos, by = c("gender", "race")) |>
  slice_sample(prop = 0.10) |>
  mutate(
    gender_label = recode(gender,
      F = "Female", M = "Male", A = "Trans Male",
      E = "Trans Female", U = "Unknown"
    )
  ) |>
  ggplot(aes(x = age, y = bmi, color = gender_label)) +
  geom_point(alpha = 0.25, size = 0.7) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.9) +
  facet_grid(rows = vars(gender_label), cols = vars(race)) +
  scale_color_brewer(palette = "Set1", name = "Sex/Gender") +
  labs(
    title    = "BMI by Age: Faceted by Sex/Gender (rows) and Race/Ethnicity (columns)",
    subtitle = "10% random sample; panels with < 50 valid observations excluded",
    x        = "Age (years)",
    y        = "BMI (kg/m²)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title      = element_text(face = "bold"),
    legend.position = "none",
    strip.text      = element_text(size = 8, face = "bold"),
    axis.text.x     = element_text(size = 7)
  )
```

---

## E) Missing Values

```{r missing-rates}
# Function to compute missingness rate for a data frame
miss_summary <- function(df, table_name) {
  df |>
    summarise(across(everything(), ~ mean(is.na(.)))) |>
    pivot_longer(everything(), names_to = "Variable", values_to = "Missing Rate") |>
    mutate(
      Table            = table_name,
      `Missing Rate`   = round(`Missing Rate` * 100, 2),
      `Missing Rate %` = paste0(`Missing Rate`, "%")
    ) |>
    select(Table, Variable, `Missing Rate %`)
}

bind_rows(
  miss_summary(patients,   "patients"),
  miss_summary(encounters, "encounters"),
  miss_summary(medications,"medications"),
  miss_summary(diseases,   "diseases")
) |>
  filter(`Missing Rate %` != "0%") |>
  kable(caption = "Missingness rates across key tables (variables with any missing)") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  scroll_box(height = "500px")
```

### Discussion

Below I discuss the variables most relevant to typical analyses, focusing on those with
notable missingness rates.

**1. Encounters – `discharge_date`**
A substantial share of encounters lack a discharge date. This is expected for outpatient
or same-day visits that were not formally discharged, but it prevents calculation of
length of stay for those records. Any inpatient-focused analysis (e.g., LOS, resource
utilization) will be restricted to encounters with a recorded discharge date, which is a
potentially non-random subset. Analysts should flag this and investigate whether
discharge-date missingness is related to department type or era.

**2. Encounters – vital signs (`bp_systolic`, `bp_diastolic`, `temperature`, `pulse`,
`height`, `weight`)**
All vital sign fields carry meaningful missingness. `height` and `weight` are missing in
a large fraction of encounters, which directly limits BMI analysis. Missing vitals are
common in administrative encounter data and may reflect outpatient visits where
measurements were not routinely recorded. Imputation would be inadvisable without
clinical context; analyses relying on vitals should use complete-case data and note the
restriction.

**3. Patients – `race`**
Race/ethnicity has notable missingness in the patient table. For health equity analyses—
a primary use case for this kind of hospital data—missing race limits stratification and
disparity estimation. Race is frequently missing non-randomly (e.g., more often missing
for certain groups), so complete-case analyses may introduce bias.

**4. Patients – `marital_status`, `occupation`, `religion`**
These optional sociodemographic fields have meaningful missingness. While less central to
clinical analyses, they are important for social determinants of health research. The
default of `"UNKNOWN"` means some apparent non-missingness may reflect the default rather
than a known value.

**5. Medications – `dose`, `dose_units`, `rate`, `rate_units`**
Dose and rate information are frequently missing. For pharmacoepidemiological analyses or
drug utilization review, this limits the ability to assess exposure levels. Presence of
a medication record without a dose may still be useful for some purposes (e.g.,
prevalence of prescribing), but any dose-response analysis would be severely hampered.

---

## F) Data Validation

### Ranges and Unreasonable Values

```{r validation-ranges}
encounters |>
  select(age, bp_systolic, bp_diastolic, temperature, pulse, height, weight) |>
  summarise(across(everything(), list(
    min    = ~ min(., na.rm = TRUE),
    max    = ~ max(., na.rm = TRUE),
    mean   = ~ round(mean(., na.rm = TRUE), 1),
    n_miss = ~ sum(is.na(.))
  ))) |>
  pivot_longer(everything(),
               names_to  = c("Variable", ".value"),
               names_sep = "_(?=[^_]+$)") |>
  kable(caption = "Range and missingness of key numeric variables in encounters") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

### Discussion

**1. Age**
The data dictionary constrains age to 0–120. Values outside this range should be treated
as errors. It is also worth checking whether patient ages at encounter are consistent
with calculated ages from birth dates (addressed in Part 2, Option 2). Ages at the very
extremes (e.g., 0 or 120) are plausible but warrant scrutiny.

**2. Blood Pressure**
Adult systolic BP is typically 70–200 mmHg and diastolic 40–130 mmHg in clinical
settings. Any values at or near zero, or extreme outliers (e.g., systolic > 250), are
likely data entry errors and should be set to missing or reviewed. A check for
physiologically implausible pulse pressure (systolic − diastolic < 5 or > 200) can also
flag erroneous pairs.

```{r bp-check}
encounters |>
  filter(!is.na(bp_systolic), !is.na(bp_diastolic)) |>
  summarise(
    n_systolic_low    = sum(bp_systolic < 50),
    n_systolic_high   = sum(bp_systolic > 250),
    n_diastolic_low   = sum(bp_diastolic < 20),
    n_diastolic_high  = sum(bp_diastolic > 150),
    n_pp_implausible  = sum((bp_systolic - bp_diastolic) < 5 |
                              (bp_systolic - bp_diastolic) > 200)
  ) |>
  pivot_longer(everything(), names_to = "Check", values_to = "Count") |>
  kable(caption = "Blood pressure plausibility checks") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

**3. Height and Weight / BMI**
Adult heights below ~48 inches (122 cm) or above ~96 inches (244 cm) are clinically
implausible for adults. Similarly, weights below ~50 lbs or above ~800 lbs in adult
populations are likely errors. These should be flagged and set to missing.

```{r bmi-extremes}
enc_pat |>
  filter(!is.na(bmi)) |>
  summarise(
    n_bmi_low    = sum(bmi < 10),
    n_bmi_high   = sum(bmi > 70),
    n_height_low = sum(height < 48, na.rm = TRUE),
    n_height_high= sum(height > 96, na.rm = TRUE),
    n_weight_low = sum(weight < 50, na.rm = TRUE),
    n_weight_high= sum(weight > 700, na.rm = TRUE)
  ) |>
  pivot_longer(everything(), names_to = "Check", values_to = "Count") |>
  kable(caption = "Height, weight, and BMI plausibility checks") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

**4. Temperature**
Normal adult body temperature is roughly 95–106 °F. Values outside this range in the
data are almost certainly recording errors.

```{r temp-check}
encounters |>
  filter(!is.na(temperature)) |>
  summarise(
    n_temp_low  = sum(temperature < 90),
    n_temp_high = sum(temperature > 110),
    min_temp    = min(temperature),
    max_temp    = max(temperature)
  ) |>
  kable(caption = "Temperature plausibility checks") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

**5. Duplicate Records**
Duplicate encounter records (same `patient_id`, `admit_date`, and `department_id`) may
indicate data entry duplication.

```{r duplicates}
dup_encounters <- encounters |>
  count(patient_id, admit_date, department_id) |>
  filter(n > 1)

cat("Number of apparent duplicate encounter combinations:",
    nrow(dup_encounters), "\n")

dup_med <- medications |>
  count(encounter_id, medication_id, patient_id, date) |>
  filter(n > 1)

cat("Number of apparent duplicate medication records:",
    nrow(dup_med), "\n")
```

**Recommendations**
Values identified as out-of-range should be set to `NA` rather than deleted so that
other fields in the same record remain usable. Definitive removal should only follow
clinical expert review. Duplicate records should be deduplicated by retaining one
instance, with documentation of the deduplication rule.

---

# Part 2

## Option 1: Latest-BMI-Only Analysis

Repeating the BMI analysis from C1, but using only the *most recent* valid height and
weight measurement per patient. "Most recent" is defined as the encounter with the
largest `admit_date` that has non-missing height and weight.

```{r latest-bmi}
latest_bmi <- encounters |>
  filter(!is.na(height), !is.na(weight), height > 0, weight > 0) |>
  mutate(bmi = (weight / height^2) * 703) |>
  left_join(patients |> select(patient_id, gender, race), by = "patient_id") |>
  mutate(
    gender = replace_na(gender, "UNKNOWN"),
    race   = replace_na(race, "UNKNOWN")
  ) |>
  group_by(patient_id) |>
  slice_max(order_by = admit_date, n = 1, with_ties = FALSE) |>
  ungroup()

cat("Number of patients with at least one valid BMI:", nrow(latest_bmi), "\n")
```

**Mean BMI by sex/gender (latest observation only):**

```{r latest-bmi-gender}
latest_bmi |>
  group_by(gender) |>
  summarise(
    n          = n(),
    mean_bmi   = round(mean(bmi, na.rm = TRUE), 2),
    median_bmi = round(median(bmi, na.rm = TRUE), 2),
    sd_bmi     = round(sd(bmi, na.rm = TRUE), 2),
    .groups    = "drop"
  ) |>
  rename(Gender = gender, N = n, `Mean BMI` = mean_bmi,
         `Median BMI` = median_bmi, `SD BMI` = sd_bmi) |>
  kable(caption = "Mean BMI by sex/gender (latest encounter per patient)") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

**Mean BMI by race/ethnicity (latest observation only):**

```{r latest-bmi-race}
latest_bmi |>
  group_by(race) |>
  summarise(
    n          = n(),
    mean_bmi   = round(mean(bmi, na.rm = TRUE), 2),
    median_bmi = round(median(bmi, na.rm = TRUE), 2),
    sd_bmi     = round(sd(bmi, na.rm = TRUE), 2),
    .groups    = "drop"
  ) |>
  arrange(desc(n)) |>
  rename(Race = race, N = n, `Mean BMI` = mean_bmi,
         `Median BMI` = median_bmi, `SD BMI` = sd_bmi) |>
  kable(caption = "Mean BMI by race/ethnicity (latest encounter per patient)") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

**Mean BMI by sex/gender and race/ethnicity (latest observation only):**

```{r latest-bmi-both}
latest_bmi |>
  group_by(gender, race) |>
  summarise(
    n        = n(),
    mean_bmi = round(mean(bmi, na.rm = TRUE), 2),
    .groups  = "drop"
  ) |>
  arrange(gender, race) |>
  rename(Gender = gender, Race = race, N = n, `Mean BMI` = mean_bmi) |>
  kable(caption = "Mean BMI by sex/gender and race/ethnicity (latest per patient)") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  scroll_box(height = "400px")
```

**Comparison with all-observations approach**

Restricting to the latest BMI per patient slightly changes the means for most groups
compared to the all-encounters analysis. This reflects the fact that repeated
measurements from the same patient are correlated; the all-encounter approach
overweights patients who appear frequently. The latest-BMI approach provides a more
patient-level snapshot and is generally preferable for cross-sectional comparisons.

---

## Option 2: Validating Encounter Ages

Each encounter records the patient's age at that visit. We can validate this by
computing the expected age from `birth_date` and `admit_date` and comparing it to the
recorded age. Only encounters with a non-missing `admit_date` are used (discharge date
is not required, per the instructions).

```{r age-validation}
age_check <- encounters |>
  filter(!is.na(age), !is.na(admit_date)) |>
  left_join(patients |> select(patient_id, birth_date), by = "patient_id") |>
  filter(!is.na(birth_date)) |>
  mutate(
    expected_age  = as.integer(interval(birth_date, admit_date) / years(1)),
    age_match     = (age == expected_age),
    age_diff      = age - expected_age
  )

n_total  <- nrow(age_check)
n_match  <- sum(age_check$age_match, na.rm = TRUE)
pct_match <- round(n_match / n_total * 100, 2)

cat("Total encounters with age and birth_date:", format(n_total, big.mark = ","), "\n")
cat("Encounters with correct age:             ", format(n_match, big.mark = ","), "\n")
cat("Percentage accurate:                     ", pct_match, "%\n")
```

```{r age-diff-dist, fig.width=8, fig.height=5}
age_check |>
  filter(abs(age_diff) <= 10) |>
  ggplot(aes(x = age_diff)) +
  geom_histogram(binwidth = 1, fill = "royalblue", color = "white", alpha = 0.85) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title    = "Distribution of Recorded Age minus Expected Age",
    subtitle = paste0(pct_match, "% of encounter ages match the birth-date-derived age"),
    x        = "Recorded Age − Expected Age (years)",
    y        = "Number of Encounters"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

```{r age-off-by-one}
age_check |>
  mutate(
    accuracy_category = case_when(
      age_match              ~ "Exact match",
      abs(age_diff) == 1     ~ "Off by 1 year",
      abs(age_diff) <= 5     ~ "Off by 2–5 years",
      TRUE                   ~ "Off by > 5 years"
    )
  ) |>
  count(accuracy_category) |>
  mutate(
    pct = round(n / sum(n) * 100, 2),
    `%` = paste0(pct, "%")
  ) |>
  arrange(desc(n)) |>
  select(accuracy_category, n, `%`) |>
  rename(`Accuracy Category` = accuracy_category, Count = n) |>
  kable(caption = "Breakdown of encounter-age accuracy") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

**Interpretation**

The encounter-level age field is largely accurate. The most common discrepancy is an
off-by-one error, which arises naturally when the encounter falls before the patient's
birthday in a given year: if a patient turns 35 in October but was admitted in January,
their recorded age might be 34 or 35 depending on how the field was populated (integer
truncation vs. rounding). This suggests the recorded age is computed by simple year
subtraction rather than accounting for whether the birthday has passed in the admit year.
Encounters with discrepancies larger than one year are likely data errors and should be
flagged for review.

---
